---
title: "BBRF GMM: Latent Daily Psychophysiological States & Suicide Risk"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
params:
  profile: "default"
editor_options: 
  markdown: 
    wrap: 72
---

```{r global, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE, comment = "")
```

```{r environment, warning = FALSE, message = FALSE, echo = FALSE}

## Load packages, source code, and config ##

suppressPackageStartupMessages({
  library(here);
  library(dplyr);
  library(readr);
  library(tidyr);
  library(ggplot2);
  library(purrr);
  library(glue);
  library(kableExtra)
})

#0.1 Source the from-scratch model code and shared utilities
source(here::here("src", "utils.R"))
source(here::here("src", "gmm_em.R"))

#0.2 Load config (single source of truth) and set the seed
cfg <- load_config(profile = params$profile)
set.seed(cfg$gmm$seed)

#0.3 Read the analytic person-day table built by R/01_build_dataset.R
daily_path <- list.files(cfg$paths$data_processed, pattern = "^ema_daily",
                         full.names = TRUE) |> sort() |> dplyr::last()

daily <- readr::read_csv(daily_path, show_col_types = FALSE)

```

## Analysis Overview

Goal is to discover latent daily states from a small set of continuous
psychophysiological features (i.e., mood, actigraphy derived sleep,
go/no-go cognition) by fitting a Gaussian Mixture Model via EM, then
characterize those states by their suicidal-ideation profiles and
analyze how individuals move among them over time.

**This will occur in three stages.**

1.  *Univariate EM (EDA + code stepping-stone).* Fit a 1-D GMM to each
    feature to inspect multimodality before committing to the joint
    model. We also run a pre-registered check on the go/no-go commission
    feature and document its eventual exclusion from the mixture
2.  *Multivariate GMM (the model).* Fit the joint mixture; select K by
    comparing BIC, AIC, and ICL (with classification entropy and the
    smallest mixing weight as degeneracy diagnostics), a Kneedle knee, a
    marginal density-fit comparison, and the person x state artifact
    check (never a single minimum)
3.  *States + dynamics.* Overlay SI/SB onto states; decode per-person
    state sequences; relate early-warning indicators (variance, lag-1
    autocorrelation) to transitions into high-SI states

```{r prepare_features, echo = FALSE, warning = FALSE}

## 1. Build the standardized feature matrix (complete cases) ##

#1.1 Configured features. The FULL set (`feat`) drives Stage-1 EDA and the
#    documented commission check; the MODELING set (`feat_model`) drops any
#    features flagged for documented exclusion (e.g., commission_elogit) so they
#    never enter the mixture, while their evidence is retained.
feat <- cfg$features
feat_excl <- if (!is.null(cfg$gmm$exclude_from_mixture)) unlist(cfg$gmm$exclude_from_mixture) else character(0)
feat_model <- setdiff(feat, feat_excl)
stopifnot(all(feat %in% names(daily)))

#1.2 Complete-case on the MODELING features (so an excluded feature's missingness
#    can't shrink N), while carrying the excluded feature(s) + overlays for
#    documentation and Stage-3 characterization.
model_df <- daily |>
  dplyr::select(subject, day, dplyr::all_of(feat),
                dplyr::any_of(c(cfg$overlay$primary, cfg$overlay$secondary, cfg$overlay$rare))) |>
  tidyr::drop_na(dplyr::all_of(feat_model))

message(glue::glue(
  "Modeling on {nrow(model_df)} complete person-days ",
  "across {dplyr::n_distinct(model_df$subject)} subjects ",
  "(of {nrow(daily)} total person-days). Modeling features: ",
  "{paste(feat_model, collapse = ', ')}."))

#1.3 Standardize the MODELING features; keep person id aligned for the artifact check
X <- zscore_matrix(model_df[, feat_model])
person_id <- model_df$subject

```

## 1. Univariate EM (per-feature multimodality)

```{r stage1_feat_dists, echo = FALSE, warning = FALSE}

## Stage 1a: feature distributions (shape QC) shown for the FULL set ##

#1a.1 Long-format the full feature set in native units, incl. commission so its
#     floor-heavy distribution is documented before the exclusion section
feat_long <- model_df |>
  dplyr::select(dplyr::all_of(feat)) |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "feature", values_to = "value") |>
  tidyr::drop_na(value)

#1a.2 Faceted histograms: read for skew, spikes/floors, and modality
ggplot(feat_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "#21908C", colour = "white", linewidth = 0.2) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  labs(title = "Feature distributions (native units)", x = NULL, y = "Person-days") +
  theme_hw()

#1a.3 Numeric shape summary, incl. share sitting exactly at the floor (the spike)
feat_long |>
  dplyr::group_by(feature) |>
  dplyr::summarise(
    n = dplyr::n(),
    mean = mean(value), sd = stats::sd(value),
    skew = mean(((value - mean(value)) / stats::sd(value))^3),
    pct_at_floor = round(100 * mean(value == min(value)), 1),
    .groups  = "drop") |>
  nice_tbl(caption = "Feature shape summary (skew + floor share)")

```

```{r stage1_univariate_em, warning = FALSE}

## Fit a 2-component 1-D GMM to each feature as exploratory structure check ##

uni_fits <- purrr::map(feat, function(v) {
  x <- model_df[[v]]
  x <- x[is.finite(x)]
  gmm_em_fit_1d(x, K = 2L, n_restarts = cfg$gmm$n_restarts, seed = cfg$gmm$seed)
})

names(uni_fits) <- feat

#1.1 Summarize each feature's recovered component means / weights
uni_summary <- purrr::imap_dfr(uni_fits, function(f, nm) {
  ord <- order(f$mu)
  tibble::tibble(
    feature = nm,
    mu_low = f$mu[ord][1], mu_high = f$mu[ord][2],
    sd_low = sqrt(f$sigma2[ord][1]), sd_high = sqrt(f$sigma2[ord][2]),
    w_low = f$weights[ord][1], w_high  = f$weights[ord][2],
    separation = abs(diff(f$mu)) / sqrt(mean(f$sigma2)))
})

nice_tbl(uni_summary, caption = "Univariate 2-component summaries", digits = 3)

```

## Commission: documented exclusion check

Go/no-go commission-error rate is a theoretically central index of
inhibitory control, and we retained it as a candidate state feature. But
\~45% of person-days have zero commission errors (\<=1 error on \~50
trials; see the floor share above), so the empirical-logit rate carries
a point mass that a Gaussian component can only absorb by collapsing its
variance. The check below fits the mixture with commission included and
shows the consequence, i.e., commission dominates feature discrimination
only by spawning near-zero-variance components anchored at the
error-rate floor. In other words, it sorts days into "any errors vs
none" rather than indexing graded control. On that pre-registered basis
we exclude it from the mixture and retain it as a descriptive overlay on
the mood/sleep states. The evidence is kept here for the writeup.

```{r commission_documented_exclusion, warning = FALSE}

## Why commission is an overlay, not a mixture input ##

if ("commission_elogit" %in% feat) {

  #C.1 Fit on commission-complete days, on the FULL feature set, at a documented K
  doc_rows <- which(stats::complete.cases(model_df[, feat]))
  X_doc <- zscore_matrix(model_df[doc_rows, feat])
  k_doc <- cfg$gmm$commission_doc_k

  fit_doc <- gmm_em_fit(X_doc, K = k_doc, n_restarts = cfg$gmm$n_restarts,
                        max_iter = cfg$gmm$max_iter, tol = cfg$gmm$tol,
                        ridge = cfg$gmm$ridge, seed = cfg$gmm$seed)

  #C.2 Commission dominates discrimination (via near-zero within-component variance)
  nice_tbl(gmm_feature_discrimination(fit_doc, X_doc),
           caption = glue::glue("Commission INCLUDED (K = {k_doc}): per-feature discrimination ",
                               "(commission's ratio is inflated by a near-zero-variance spike)"))

  #C.3 The smoking gun: states whose commission var_z sits at the ridge are the
  #    degenerate zero-error spike(s)
  ci <- which(colnames(X_doc) == "commission_elogit")
  nice_tbl(tibble::tibble(
    state = factor(seq_len(fit_doc$K)),
    mean_z = round(fit_doc$mu[, ci], 2),
    var_z  = round(vapply(fit_doc$Sigma, function(S) S[ci, ci], numeric(1)), 4),
    weight = round(fit_doc$weights, 3)),
    caption = "commission_elogit by state (var_z near the GMM ridge => degenerate spike)")

} else {
  message("commission_elogit not in the configured feature set; skipping documented check.")
}

```

## 2. Multivariate GMM + model selection

```{r gmm_select_k, warning = FALSE}

## Fit across K; compare BIC, AIC, ICL (+ entropy, smallest mixing weight) ##

sel <- gmm_select_k(
  X,
  k_range = unlist(cfg$gmm$k_range),
  n_restarts = cfg$gmm$n_restarts,
  max_iter = cfg$gmm$max_iter,
  tol = cfg$gmm$tol,
  ridge = cfg$gmm$ridge,
  seed = cfg$gmm$seed)

nice_tbl(sel$bic_table,
         caption = "Model selection across K (BIC, AIC, ICL, entropy, min mixing weight)",
         digits = 1)

#2.1 Objective knees on the BIC and ICL curves
k_knee_bic <- kneedle(sel$bic_table$K, sel$bic_table$bic)
k_knee_icl <- kneedle(sel$bic_table$K, sel$bic_table$icl)

message(glue::glue("BIC min K={sel$best_k} | ICL min K={sel$best_k_icl} | ",
                   "Kneedle(BIC) K={k_knee_bic} | Kneedle(ICL) K={k_knee_icl}"))

#2.2 BIC vs ICL: ICL adds an overlap/entropy penalty, so it resists the
#    over-splitting BIC shows when the mixture tiles density rather than finding
#    distinct subgroups
crit_long <- sel$bic_table |>
  dplyr::select(K, BIC = bic, ICL = icl) |>
  tidyr::pivot_longer(c(BIC, ICL), names_to = "criterion", values_to = "value")

ggplot(crit_long, aes(K, value, colour = criterion)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.4) +
  geom_vline(xintercept = sel$best_k_icl, linetype = 2, colour = "grey50") +
  scale_colour_viridis_d(end = 0.6) +
  scale_x_continuous(breaks = sel$bic_table$K) +
  labs(title = "Model selection: BIC vs ICL",
       subtitle = glue::glue("dashed = ICL-min (K={sel$best_k_icl}); ",
                             "Kneedle(ICL) K={k_knee_icl}; lower is better"),
       x = "Number of states (K)", y = "Criterion") +
  theme_hw()

#2.3 Candidate K set for the density comparison (DYNAMIC; config-overridable).
#    Defaults to the ICL minimum, the ICL knee, and K=2 if not specified given
#    K=2 relevance derived from viewing the density plots
candidate_k <- if (!is.null(cfg$gmm$candidate_k)) {
  sort(unique(unlist(cfg$gmm$candidate_k)))
} else {
  sort(unique(c(2L, sel$best_k_icl, k_knee_icl)))
}

```

```{r gmm_choose_k, warning = FALSE, echo = FALSE}

## 2.4 Lock K from all lenses (config override, else the ICL minimum) ##
chosen_k <- if (!is.null(cfg$gmm$chosen_k)) cfg$gmm$chosen_k else sel$best_k_icl
best <- sel$fits[[as.character(chosen_k)]]
message(glue::glue(
  "Carrying K = {chosen_k} ",
  "({ifelse(is.null(cfg$gmm$chosen_k), 'ICL default', 'config override')}). ",
  "Set cfg$gmm$chosen_k in config to lock it after reviewing the criteria table + curve."))

```

```{r gmm_density_compare, fig.height = 7, warning = FALSE}

## 2.4 What each state's distribution looks like (at the chosen K) ##

#2.4.1 PLOT A: per-feature component densities, one curve PER STATE, over the
#      data histogram. Reads as: "for this feature, here is each state's density
#      and where it sits." The summed (mixture) density is the dashed envelope
comp_dens <- purrr::map_dfr(colnames(X),
  function(f) gmm_component_density(best, X, f))

mix_dens  <- purrr::map_dfr(colnames(X),
  function(f) gmm_marginal_density(best, X, f))

model_long <- model_df |>
  dplyr::select(dplyr::all_of(colnames(X))) |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "feature", values_to = "value")

ggplot() +
  geom_histogram(data = model_long, aes(value, y = ggplot2::after_stat(density)),
                 bins = 30, fill = "grey88", colour = "white", linewidth = 0.2) +
  geom_area(data = comp_dens, aes(value, density, fill = state),
            position = "identity", alpha = 0.35) +
  geom_line(data = comp_dens, aes(value, density, colour = state), linewidth = 0.7) +
  geom_line(data = mix_dens, aes(value, density), colour = "grey25",
            linetype = 2, linewidth = 0.5) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  scale_fill_viridis_d(end = 0.9) + scale_colour_viridis_d(end = 0.9) +
  labs(title = glue::glue("State distributions by feature (K = {best$K})"),
       subtitle = "Each colour = one state's weighted Gaussian; dashed grey = full mixture",
       x = NULL, y = "Density") +
  theme_hw()

#2.4.2 PLOT B: the mood plane, coloured by hard state assignment, with each
#      state's mean marked. Depression x anxiety = the dominant axis
if (all(c("depression", "anxiety") %in% colnames(X))) {
  state_means <- gmm_state_profile(best, X)

  ggplot(model_df, aes(depression, anxiety)) +
    geom_point(aes(colour = factor(gmm_map_assign(best))), alpha = 0.5, size = 1.4) +
    geom_point(data = state_means, aes(depression, anxiety),
               shape = 21, size = 5, stroke = 1.1, fill = "white", colour = "grey20") +
    geom_text(data = state_means, aes(depression, anxiety, label = state),
              size = 3.2, fontface = "bold") +
    scale_colour_viridis_d(end = 0.9, name = "State") +
    labs(title = glue::glue("Person-days on the mood plane, by state (K = {best$K})"),
         subtitle = "Points = person-days (coloured by MAP state); numbered rings = state means",
         x = "Depression", y = "Anxiety") +
    theme_hw()
}

#2.4.3 PLOT C: standardized state x feature heatmap (compact fingerprint).
#      A single grid: rows = states, columns = features, fill = z-mean. Fast read
#      of "what is high/low in each state" across ALL features at once
fing_long <- best$mu |>
  `colnames<-`(colnames(X)) |>
  tibble::as_tibble() |>
  dplyr::mutate(state = factor(seq_len(best$K))) |>
  tidyr::pivot_longer(-state, names_to = "feature", values_to = "z_mean")

ggplot(fing_long, aes(feature, state, fill = z_mean)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%+.1f", z_mean)), size = 3) +
  scale_fill_gradient2(low = "#3B528B", mid = "grey95", high = "#B8003E",
                       midpoint = 0, name = "z-mean") +
  labs(title = glue::glue("State fingerprints (standardized; K = {best$K})"),
       subtitle = "Fill = standardized mean (0 = sample average); +red above, -blue below",
       x = NULL, y = "State") +
  theme_hw() + theme(axis.text.x = element_text(angle = 30, hjust = 1))

```

```{r gmm_state_profiles, warning = FALSE}

## 2.6 What ARE the states? Native-unit profiles + standardized fingerprints ##

nice_tbl(gmm_state_profile(best, X), digits = 2,
         caption = "Latent state profiles (native units) and state size")

#2.6.1 Standardized fingerprint (z-units; 0 = sample average), lollipop per state
fing <- best$mu

colnames(fing) <- colnames(X)

fing <- tibble::as_tibble(fing)

fing$state <- factor(seq_len(best$K))

fing_long <- tidyr::pivot_longer(fing, -state, names_to = "feature", values_to = "z_mean")

ggplot(fing_long, aes(x = z_mean, y = feature, colour = state)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey60") +
  geom_segment(aes(x = 0, xend = z_mean, yend = feature), colour = "grey80", linewidth = 0.6) +
  geom_point(size = 3) +
  facet_wrap(~ state, nrow = 1) +
  scale_colour_viridis_d(end = 0.9) +
  labs(title = "Latent state fingerprints",
       subtitle = "Standardized feature means by state (0 = sample average)",
       x = "Standardized mean (SD units)", y = NULL) +
  theme_hw() + theme(legend.position = "none")

```

```{r gmm_artifact_check, warning = FALSE}

## 2.7 Person x state check: is any state just one person's days? ##
chk <- gmm_person_state_check(best, person_id)

nice_tbl(data.frame(state = seq_along(chk$max_person_share),
                    n_days = as.integer(chk$state_totals),
                    max_single_person_share = round(chk$max_person_share, 3)),
         caption = "Per-state size and single-person dominance (high share => trait artifact)")

```

## 2.8 Sensitivity: within-person (person-mean-centered) refit

The primary model is fit on raw (across-sample standardized) person-days, so a
fair objection is that the states might be sorting *people* (chronically
high-distress vs chronically euthymic individuals) rather than capturing
*within-person day-to-day shifts* -- the latter being the dynamical-systems
claim we actually want to make. The artifact check already shows no state is one
person's days; this sensitivity goes further by asking whether the same state
structure survives when **all** stable between-person differences are removed.

Two reads. First, the per-feature **ICC(1)** quantifies how much of each
feature's variance is between-person (trait) vs within-person (state); the lower
the ICC, the more a feature reflects genuine day-to-day movement, and the more
defensible the uncentered model is as the primary. Second, we refit the entire
GMM on **person-mean-centered** features (each day expressed relative to that
person's own average) and ask whether (a) model selection still prefers a
comparable K, (b) the centered states show the same qualitative fingerprints,
and (c) the day-level assignments agree with the primary solution (adjusted Rand
index). We do not expect *identical* partitions -- centering deliberately moves a
chronically-depressed person's average day off the high-distress state -- so the
scientific question is whether a recognizable distress/euthymic structure
*re-emerges from purely within-person variation*, which is exactly the claim the
dynamical framing needs.

```{r centered_icc, warning = FALSE}

## 2.8.1 Per-feature ICC(1): between-person (trait) vs within-person (state) ##

#2.8.1 One-way random-effects ICC via base-R aov (no new dependencies). High ICC
#      => the feature mostly separates people (trait-like); low ICC => it mostly
#      tracks within-person change (state-like, the signal we want). k0 is the
#      unbalanced-design average group size.
icc1 <- function(value, group) {
  group <- factor(group)
  ni <- table(group); g <- length(ni); N <- length(value)
  ms  <- summary(stats::aov(value ~ group))[[1]][, "Mean Sq"]
  msb <- ms[1]; msw <- ms[2]
  k0  <- (N - sum(ni^2) / N) / (g - 1)
  max((msb - msw) / (msb + (k0 - 1) * msw), 0)         # floor at 0
}

icc_tbl <- tibble::tibble(
  feature = feat_model,
  icc = vapply(feat_model, function(f) icc1(model_df[[f]], person_id), numeric(1))) |>
  dplyr::mutate(pct_within = round(100 * (1 - icc), 1), icc = round(icc, 3)) |>
  dplyr::arrange(icc)

nice_tbl(icc_tbl,
         caption = "Per-feature ICC(1): trait (between) vs state (within) variance share")

```

```{r centered_refit, fig.height = 4.5, warning = FALSE}

## 2.8.2 Refit the GMM on person-mean-centered (within-person) features ##

#2.8.2.1 Strip each person's own mean from every modeling feature, then
#        re-standardize so the centered matrix carries the center/scale
#        attributes the profilers need. Centered features encode "this day vs
#        this person's average," removing all stable between-person differences.
Xc <- zscore_matrix(person_center(model_df[, feat_model], person_id))

#2.8.2.2 Run the SAME selection machinery on the centered data: does centering
#        change the preferred K? (structure stability across the four lenses)
sel_c <- gmm_select_k(
  Xc, k_range = unlist(cfg$gmm$k_range), n_restarts = cfg$gmm$n_restarts,
  max_iter = cfg$gmm$max_iter, tol = cfg$gmm$tol, ridge = cfg$gmm$ridge,
  seed = cfg$gmm$seed)

k_knee_bic_c <- kneedle(sel_c$bic_table$K, sel_c$bic_table$bic)
k_knee_icl_c <- kneedle(sel_c$bic_table$K, sel_c$bic_table$icl)
message(glue::glue(
  "CENTERED selection: BIC min K={sel_c$best_k} | ICL min K={sel_c$best_k_icl} | ",
  "Kneedle(BIC) K={k_knee_bic_c} | Kneedle(ICL) K={k_knee_icl_c}"))

#2.8.2.3 Hold K at the PRIMARY chosen_k so the two partitions are directly
#        comparable, then read the centered states' standardized fingerprints
best_c <- sel_c$fits[[as.character(chosen_k)]]

fing_c <- best_c$mu
colnames(fing_c) <- colnames(Xc)
fing_c <- tibble::as_tibble(fing_c)
fing_c$state <- factor(seq_len(best_c$K))
fing_c_long <- tidyr::pivot_longer(fing_c, -state, names_to = "feature", values_to = "z_mean")

ggplot(fing_c_long, aes(feature, state, fill = z_mean)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%+.1f", z_mean)), size = 3) +
  scale_fill_gradient2(low = "#3B528B", mid = "grey95", high = "#B8003E",
                       midpoint = 0, name = "z-mean") +
  labs(title = glue::glue("Within-person state fingerprints (centered; K = {best_c$K})"),
       subtitle = "Person-mean-centered features; compare structure to the primary fit",
       x = NULL, y = "State") +
  theme_hw() + theme(axis.text.x = element_text(angle = 30, hjust = 1))

```

```{r centered_agreement, warning = FALSE}

## 2.8.3 Do the two solutions agree? Confusion matrix + adjusted Rand index ##

#2.8.3.1 Adjusted Rand index from scratch: partition agreement corrected for
#        chance (0 = chance, 1 = identical). Invariant to label permutation, so
#        it sidesteps GMM label-switching between the two fits.
adjusted_rand <- function(a, b) {
  tab <- table(a, b); cc <- function(x) choose(x, 2)
  sij <- sum(cc(tab)); si <- sum(cc(rowSums(tab))); sj <- sum(cc(colSums(tab)))
  expected <- si * sj / cc(sum(tab)); maxidx <- (si + sj) / 2
  (sij - expected) / (maxidx - expected)
}

state_primary  <- gmm_map_assign(best)
state_centered <- gmm_map_assign(best_c)
ari <- adjusted_rand(state_primary, state_centered)

#2.8.3.2 Confusion matrix: rows = primary (uncentered) state, cols = centered
xtab <- as.data.frame.matrix(table(primary = state_primary, centered = state_centered))
xtab <- tibble::rownames_to_column(xtab, "primary \\ centered")
nice_tbl(xtab, digits = 0,
         caption = glue::glue("Primary x centered assignment (adjusted Rand index = {round(ari, 3)})"))

#2.8.3.3 Does the high-SI gradient survive centering? SI overlay by CENTERED state
model_df$state_centered <- factor(state_centered)
overlay_vars_c <- intersect(c(cfg$overlay$primary, cfg$overlay$secondary, cfg$overlay$rare),
                            names(model_df))
nice_tbl(
  model_df |>
    dplyr::group_by(state_centered) |>
    dplyr::summarise(n_days = dplyr::n(),
                     dplyr::across(dplyr::all_of(overlay_vars_c), ~ mean(.x, na.rm = TRUE)),
                     .groups = "drop"),
  digits = 3, caption = "SI/SB profile by within-person (centered) state")

```

## 2.9 EM verification & density-estimation adequacy

The mixture is hand-implemented, so before any scientific claim we verify the
*algorithm* and the *density estimate* separately. Four checks, each targeting a
distinct failure mode:

1. **EM monotonicity** -- EM provably never decreases the observed-data
   log-likelihood; a single decrease in the iteration trace means the
   implementation is wrong. This is the sharpest single correctness test
   available without a reference library.
2. **Restart stability** -- EM only finds local optima. If the 30 restarts
   scatter across many distinct log-likelihood values, the reported solution is
   initialization luck; if most restarts reach the same best mode, the optimum
   is reliable.
3. **Posterior predictive check (PPC)** -- a good *density estimate* must be
   able to regenerate the data. We simulate person-days from the fitted mixture
   and compare simulated vs observed marginals and the full correlation
   structure (the joint dependence is what a multivariate GMM buys over per-
   feature fits). PIT values (the fitted marginal CDF evaluated at each
   observation) should be approximately Uniform(0,1) if the marginals fit.
4. **Person-blocked cross-validated log-likelihood** -- the direct, assumption-
   free answer to "BIC's penalty assumes independent rows." Held-out density
   (fit on most people, score the held-out people's days) makes no effective-N
   assumption: if K = 4 generalizes to unseen *people*, its held-out
   log-likelihood should be competitive, and overfit K should drop.

```{r em_verification, fig.height = 4, warning = FALSE}

## 2.9.1 EM correctness: monotone log-likelihood + restart stability ##

#2.9.1.1 Monotonicity audit on the chosen fit's iteration trace. Tolerance 1e-8
#        absorbs floating-point noise only; any real decrease fails loudly.
ll_tr <- best$log_lik_trace
mono_viol <- sum(diff(ll_tr) < -1e-8)
message(glue::glue(
  "EM monotonicity: {length(ll_tr)} iterations, {mono_viol} decreases ",
  "(must be 0); total log-lik climb = {round(ll_tr[length(ll_tr)] - ll_tr[1], 1)}."))

ggplot(data.frame(iter = seq_along(ll_tr), loglik = ll_tr), aes(iter, loglik)) +
  geom_line(colour = "#21908C", linewidth = 0.7) +
  labs(title = glue::glue("EM convergence trace (K = {best$K})"),
       subtitle = "Monotone non-decreasing by construction; any dip = implementation bug",
       x = "EM iteration", y = "Observed-data log-likelihood") +
  theme_hw()

#2.9.1.2 Restart stability: final log-lik of every restart at the chosen K.
#        "Modal share" = fraction of restarts landing within 1e-3 of the best.
rl <- best$restart_loglik[is.finite(best$restart_loglik)]
modal_share <- mean(rl > max(rl) - 1e-3)
message(glue::glue(
  "Restart stability: {length(rl)} restarts, {round(100 * modal_share, 1)}% ",
  "reached the best mode (spread = {round(max(rl) - min(rl), 1)} log-lik units)."))

ggplot(data.frame(loglik = rl), aes(x = loglik, y = "")) +
  geom_jitter(height = 0.15, size = 2, alpha = 0.6, colour = "grey40") +
  geom_point(aes(x = max(rl), y = ""), colour = "#21908C", size = 3.5) +
  labs(title = glue::glue("Final log-likelihood across {length(rl)} restarts (K = {best$K})"),
       subtitle = "Teal = retained solution; tight clustering = reliable optimum",
       x = "Final log-likelihood", y = NULL) +
  theme_hw()

```

```{r density_ppc, fig.height = 6, warning = FALSE}

## 2.9.2 Posterior predictive check: can the fitted mixture regenerate the data? ##

#2.9.2.1 Simulate as many person-days as observed, de-standardize to native units
sim <- gmm_simulate(best, n = nrow(X), seed = cfg$gmm$seed)
sim_native <- sweep(sweep(sim$X, 2L, attr(X, "scale"), "*"), 2L, attr(X, "center"), "+")

ppc_long <- dplyr::bind_rows(
  model_df |> dplyr::select(dplyr::all_of(colnames(X))) |>
    tidyr::pivot_longer(dplyr::everything(), names_to = "feature", values_to = "value") |>
    dplyr::mutate(source = "observed"),
  tibble::as_tibble(sim_native) |>
    tidyr::pivot_longer(dplyr::everything(), names_to = "feature", values_to = "value") |>
    dplyr::mutate(source = "simulated"))

#2.9.2.2 Overlay: simulated density should trace the observed histogram. Misfit
#        shows as systematic gaps (e.g., unmodeled skew or floors).
ggplot() +
  geom_histogram(data = dplyr::filter(ppc_long, source == "observed"),
                 aes(value, y = ggplot2::after_stat(density)),
                 bins = 30, fill = "grey85", colour = "white", linewidth = 0.2) +
  geom_density(data = dplyr::filter(ppc_long, source == "simulated"),
               aes(value), colour = "#21908C", linewidth = 0.8) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  labs(title = glue::glue("Posterior predictive check (K = {best$K})"),
       subtitle = "Grey = observed person-days; teal = density of mixture-simulated person-days",
       x = NULL, y = "Density") +
  theme_hw()

#2.9.2.3 Joint structure: the simulated CORRELATION matrix must match the
#        observed one (this is what the full covariance components buy). Report
#        the worst absolute discrepancy across all feature pairs.
cor_obs <- stats::cor(X)
cor_sim <- stats::cor(sim$X)
cor_diff <- cor_sim - cor_obs
message(glue::glue(
  "PPC correlation recovery: max |simulated - observed| correlation = ",
  "{round(max(abs(cor_diff[upper.tri(cor_diff)])), 3)} across ",
  "{sum(upper.tri(cor_diff))} feature pairs."))

cor_long <- dplyr::bind_rows(
  tibble::as_tibble(cor_obs, rownames = "f1") |>
    tidyr::pivot_longer(-f1, names_to = "f2", values_to = "r") |>
    dplyr::mutate(source = "observed"),
  tibble::as_tibble(cor_sim, rownames = "f1") |>
    tidyr::pivot_longer(-f1, names_to = "f2", values_to = "r") |>
    dplyr::mutate(source = "simulated"))

ggplot(cor_long, aes(f2, f1, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%+.2f", r)), size = 2.6) +
  facet_wrap(~ source) +
  scale_fill_gradient2(low = "#3B528B", mid = "grey95", high = "#B8003E",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  labs(title = "Joint-structure recovery: observed vs simulated correlations",
       subtitle = "A faithful mixture reproduces the dependence structure, not just the marginals",
       x = NULL, y = NULL) +
  theme_hw() + theme(axis.text.x = element_text(angle = 35, hjust = 1, size = 7),
                     axis.text.y = element_text(size = 7))

```

```{r density_pit, fig.height = 3.8, warning = FALSE}

## 2.9.3 PIT: probability integral transform of each marginal ##

#2.9.3.1 For feature j, the fitted marginal CDF is sum_k w_k * pnorm(x; mu_kj,
#        sd_kj) (computed in z-space, where the fit lives). If the marginal
#        density is right, the CDF evaluated at the data is ~ Uniform(0, 1):
#        flat PIT histogram. Humps/edge spikes localize misfit.
pit_long <- purrr::map_dfr(seq_len(ncol(X)), function(j) {
  pit <- rowSums(vapply(seq_len(best$K), function(k)
    best$weights[k] * stats::pnorm(X[, j], best$mu[k, j],
                                   sqrt(best$Sigma[[k]][j, j])),
    numeric(nrow(X))))
  ks <- suppressWarnings(stats::ks.test(pit, "punif")$statistic)
  tibble::tibble(feature = colnames(X)[j], pit = pit, ks = round(ks, 3))
})

nice_tbl(dplyr::distinct(pit_long, feature, ks) |> dplyr::arrange(ks),
         caption = "PIT uniformity per feature (KS distance from Uniform(0,1); smaller = better marginal fit)")

ggplot(pit_long, aes(pit)) +
  geom_histogram(breaks = seq(0, 1, by = 0.1), fill = "#21908C",
                 colour = "white", linewidth = 0.2) +
  geom_hline(yintercept = nrow(X) / 10, linetype = 2, colour = "grey50") +
  facet_wrap(~ feature, ncol = 3) +
  labs(title = "PIT histograms (fitted marginal CDF at each observation)",
       subtitle = "Dashed = perfect-uniform reference; flat bars = well-calibrated marginal density",
       x = "PIT value", y = "Person-days") +
  theme_hw()

```

```{r cv_heldout_loglik, fig.height = 4, warning = FALSE}

## 2.9.4 Person-blocked cross-validated log-likelihood across K ##

#2.9.4.1 5 folds of SUBJECTS (never split a person across train/test), so the
#        score asks: does the K-state density generalize to unseen PEOPLE? This
#        is the independence-assumption-free counterpart to BIC. Standardization
#        is re-estimated within each training fold and projected onto the test
#        fold (no leakage).
set.seed(cfg$gmm$seed)
subj_all <- unique(person_id)
folds <- split(sample(subj_all), rep_len(1:5, length(subj_all)))
k_cv <- 1:6
n_restarts_cv <- 10L   # fewer restarts per fold to bound runtime; stability ok per 2.9.1

M_raw <- as.matrix(model_df[, feat_model])
cv_rows <- list(); i <- 0L
for (K in k_cv) {
  for (f in seq_along(folds)) {
    te_idx <- person_id %in% folds[[f]]
    Ztr <- zscore_matrix(M_raw[!te_idx, , drop = FALSE])
    Zte <- sweep(sweep(M_raw[te_idx, , drop = FALSE], 2L, attr(Ztr, "center"), "-"),
                 2L, attr(Ztr, "scale"), "/")
    fit_f <- gmm_em_fit(Ztr, K, n_restarts = n_restarts_cv,
                        max_iter = cfg$gmm$max_iter, tol = cfg$gmm$tol,
                        ridge = cfg$gmm$ridge, seed = cfg$gmm$seed + 100L * K + f)
    i <- i + 1L
    cv_rows[[i]] <- data.frame(K = K, fold = f, n_test = sum(te_idx),
                               ll = gmm_loglik(fit_f, Zte))
  }
}

#2.9.4.2 Mean held-out log-likelihood PER PERSON-DAY (comparable across folds of
#        different sizes), with across-fold SE
cv <- do.call(rbind, cv_rows) |>
  dplyr::mutate(ll_per_day = ll / n_test) |>
  dplyr::group_by(K) |>
  dplyr::summarise(mean_ll = mean(ll_per_day),
                   se_ll = stats::sd(ll_per_day) / sqrt(dplyr::n()),
                   .groups = "drop")

k_cv_best <- cv$K[which.max(cv$mean_ll)]
message(glue::glue(
  "Person-blocked CV: held-out log-lik/day peaks at K = {k_cv_best} ",
  "(chosen K = {chosen_k})."))

nice_tbl(cv, digits = 3,
         caption = "Held-out log-likelihood per person-day by K (5-fold, person-blocked)")

ggplot(cv, aes(K, mean_ll)) +
  geom_line(colour = "grey70", linewidth = 0.6) +
  geom_errorbar(aes(ymin = mean_ll - se_ll, ymax = mean_ll + se_ll),
                width = 0.12, colour = "grey55") +
  geom_point(size = 2.4, colour = "grey45") +
  geom_point(data = dplyr::filter(cv, K == chosen_k), size = 4, colour = "#21908C") +
  scale_x_continuous(breaks = k_cv) +
  labs(title = "Person-blocked cross-validated density fit",
       subtitle = glue::glue("Higher is better; teal = chosen K = {chosen_k}; ",
                             "no independence assumption (unlike BIC)"),
       x = "Number of states (K)", y = "Held-out log-lik / person-day") +
  theme_hw()

```

## 3. State characterization, SI overlay, and dynamics

```{r state_overlay, warning = FALSE}

## 3.1 MAP-assign states and overlay SI prevalence (+ commission) per state ##

model_df$state <- factor(gmm_map_assign(best))

overlay_vars <- intersect(c(cfg$overlay$primary, cfg$overlay$secondary, cfg$overlay$rare),
                          names(model_df))
state_overlay <- model_df |>
  dplyr::group_by(state) |>
  dplyr::summarise(
    n_days = dplyr::n(),
    dplyr::across(dplyr::all_of(overlay_vars), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop")

nice_tbl(state_overlay, digits = 3, caption = "SI/SB (and commission overlay) profile by latent state")

```

```{r state_sequences, fig.height = 5, warning = FALSE}

## 3.2 Stage 3a: per-person state sequences, dwell times, transitions ##

#3.2.1 Decode each subject's MAP-state sequence ordered by study day. `day` is a
#      1-30 study-day index, so consecutive days differ by exactly 1; a larger
#      gap is a missed EMA day and must NOT be read as a transition.
seq_df <- model_df |>
  dplyr::mutate(state = as.integer(as.character(state))) |>
  dplyr::arrange(subject, day) |>
  dplyr::group_by(subject) |>
  dplyr::mutate(prev_state = dplyr::lag(state),
                prev_day   = dplyr::lag(day),
                consecutive = !is.na(prev_day) & (day - prev_day == 1L)) |>
  dplyr::ungroup()

#3.2.2 Person x day state raster: each row a subject, each column a study day,
#      coloured by MAP state (blank = missing day). The raw trajectory material.
ggplot(seq_df, aes(day, factor(subject), fill = factor(state))) +
  geom_tile(colour = "grey95", linewidth = 0.2) +
  scale_fill_viridis_d(end = 0.9, name = "State") +
  labs(title = "Per-person daily state sequences",
       subtitle = "Each row = one subject across study days; gaps = missing EMA days",
       x = "Study day", y = NULL) +
  theme_hw() + theme(axis.text.y = element_text(size = 6))

#3.2.2b Assignment confidence: mean max-responsibility per state. Values near 1
#       mean confident hard assignments (well-separated states); values near 1/K
#       mean fuzzy assignments, in which case rapid day-to-day "switching" is
#       partly soft-assignment noise rather than true state change.
conf_df <- tibble::tibble(state = factor(gmm_map_assign(best)),
                          max_resp = gmm_max_resp(best)) |>
  dplyr::group_by(state) |>
  dplyr::summarise(n_days = dplyr::n(),
                   mean_max_resp = round(mean(max_resp), 3),
                   pct_confident = round(100 * mean(max_resp > 0.8), 1),
                   .groups = "drop")

nice_tbl(conf_df,
         caption = glue::glue("MAP-assignment confidence per state (1/K = {round(1/best$K, 2)} ",
                              "= chance; >0.8 = confident)"))

```

```{r state_si_overlay_raster, fig.height = 6.5, warning = FALSE}

## 3.2b SI overlaid on the per-person state raster ##

#3.2b.1 Two separate figures so neither encoding is squeezed: (A) the primary
#       CONTINUOUS metric (si_bother) as a size/opacity dot overlay; (B) the
#       BINARY events on one panel, distinguished by SHAPE. Deliberately omitted:
#       si_severity (near-collinear with si_bother -- same construct, coarser
#       scale) and ACTIVESIWITHINTENT (~constant near 1 across all states in the
#       overlay table, so it carries no discriminating signal here).
si_cont_metric <- "si_bother"
si_event_vars  <- intersect(c("si_any", "SELFHARM", "SUICIDEATTEMPT"), names(seq_df))

#3.2b.2 FIGURE A: state raster + continuous SI dots. Tile = MAP state
#       (desaturated so the overlay reads); dot size AND opacity scale with the
#       day's si_bother, so high-SI days pop and zero-SI days vanish.
ggplot(seq_df, aes(day, factor(subject))) +
  geom_tile(aes(fill = factor(state)), colour = "grey95",
            linewidth = 0.15, alpha = 0.55) +
  geom_point(data = dplyr::filter(seq_df, !is.na(.data[[si_cont_metric]]),
                                  .data[[si_cont_metric]] > 0),
             aes(size = .data[[si_cont_metric]], alpha = .data[[si_cont_metric]]),
             shape = 16, colour = "grey10") +
  scale_fill_viridis_d(end = 0.9, name = "State") +
  scale_size_continuous(range = c(0.5, 3.4), name = si_cont_metric) +
  scale_alpha_continuous(range = c(0.25, 0.95), guide = "none") +
  labs(title = "Continuous SI overlaid on daily state sequences",
       subtitle = glue::glue("Tile = MAP state; dot size/opacity = {si_cont_metric} (0 days unmarked)"),
       x = "Study day", y = NULL) +
  theme_hw() + theme(axis.text.y = element_text(size = 6))

```

```{r state_si_events_raster, fig.height = 6.5, warning = FALSE}

#3.2b.3 FIGURE B: binary suicidality EVENTS on the same raster, one panel,
#       distinguished by shape (legend). Plotted rarest-last so attempts/self-
#       harm sit on top of the (frequent) si_any marks.
events_long <- seq_df |>
  dplyr::select(subject, day, state, dplyr::all_of(si_event_vars)) |>
  tidyr::pivot_longer(dplyr::all_of(si_event_vars),
                      names_to = "event", values_to = "v") |>
  dplyr::filter(!is.na(v), v > 0) |>
  dplyr::mutate(event = factor(event, levels = c("si_any", "SELFHARM", "SUICIDEATTEMPT")))

ggplot(seq_df, aes(day, factor(subject))) +
  geom_tile(aes(fill = factor(state)), colour = "grey95",
            linewidth = 0.15, alpha = 0.55) +
  geom_point(data = events_long, aes(shape = event), size = 1.7,
             stroke = 0.7, colour = "grey10") +
  scale_fill_viridis_d(end = 0.9, name = "State") +
  scale_shape_manual(values = c(si_any = 1, SELFHARM = 17, SUICIDEATTEMPT = 4),
                     name = "Event", drop = FALSE) +
  labs(title = "Suicidality events overlaid on daily state sequences",
       subtitle = "Circle = any SI; triangle = self-harm; cross = suicide attempt",
       x = "Study day", y = NULL) +
  theme_hw() + theme(axis.text.y = element_text(size = 6))

```

```{r state_transition_network, fig.height = 6, warning = FALSE}

## 3.2c State-transition structure: network + exact-probability heatmap ##

#3.2.3 Transition structure over CONSECUTIVE-day pairs only (row = from, col = to).
#      Row-normalized => empirical transition probabilities; the diagonal is
#      state persistence (a slow, "sticky" state has high self-transition).
trans <- seq_df |>
  dplyr::filter(consecutive) |>
  dplyr::count(from = prev_state, to = state) |>
  tidyr::complete(from = seq_len(best$K), to = seq_len(best$K), fill = list(n = 0L)) |>
  dplyr::group_by(from) |>
  dplyr::mutate(prob = n / sum(n)) |>
  dplyr::ungroup()

#3.2.3a Network layout: lay the K states on a circle. Node SIZE = how often the
#       state is occupied; node FILL = that state's mean SI, so the high-risk
#       state is visually obvious; directed curved EDGES = between-state
#       transition probability (self-transitions annotated separately below).
si_metric_net <- if ("si_bother" %in% names(model_df)) "si_bother" else cfg$overlay$primary[1]
state_i_all <- as.integer(as.character(model_df$state))
ang <- (seq_len(best$K) - 1) / best$K * 2 * pi + pi / 2     # start at top
nodes <- tibble::tibble(
  state = seq_len(best$K),
  x = cos(ang), y = sin(ang),
  occupancy = as.numeric(table(factor(state_i_all, levels = seq_len(best$K)))),
  si = vapply(seq_len(best$K), function(s)
    mean(model_df[[si_metric_net]][state_i_all == s], na.rm = TRUE), numeric(1)))

#3.2.3b Off-diagonal edges only (between-state moves); drop near-zero edges so the
#       graph isn't a hairball. A perpendicular midpoint offset bends each edge so
#       opposing arrows (i->j vs j->i) don't overlap.
edges <- trans |>
  dplyr::filter(from != to, prob > 0.02) |>
  dplyr::left_join(dplyr::rename(nodes, from = state, x0 = x, y0 = y), by = "from") |>
  dplyr::left_join(dplyr::rename(nodes, to   = state, x1 = x, y1 = y), by = "to") |>
  dplyr::mutate(
    mx = (x0 + x1) / 2, my = (y0 + y1) / 2,
    nx = -(y1 - y0), ny = (x1 - x0), len = sqrt(nx^2 + ny^2),
    cx = mx + 0.12 * nx / len, cy = my + 0.12 * ny / len,   # bend control point
    a0 = atan2(cy - y0, cx - x0), a1 = atan2(cy - y1, cx - x1),
    x0t = x0 + 0.16 * cos(a0), y0t = y0 + 0.16 * sin(a0),   # trim off node centers
    x1t = x1 + 0.16 * cos(a1), y1t = y1 + 0.16 * sin(a1))

#3.2.3c Quadratic-Bezier polylines for each curved edge (base R; no igraph dep)
bez <- function(p0, p1, p2, n = 30) {
  t <- seq(0, 1, length.out = n)
  (1 - t)^2 * p0 + 2 * (1 - t) * t * p1 + t^2 * p2
}
edge_paths <- if (nrow(edges) > 0) do.call(rbind, lapply(seq_len(nrow(edges)), function(i) {
  e <- edges[i, ]
  data.frame(id = i, prob = e$prob,
             x = bez(e$x0t, e$cx, e$x1t), y = bez(e$y0t, e$cy, e$y1t))
})) else NULL

#3.2.3c.1 arrowhead = last segment of each edge path
heads <- if (!is.null(edge_paths)) edge_paths |>
  dplyr::group_by(id) |>
  dplyr::slice_tail(n = 2) |>
  dplyr::summarise(prob = dplyr::first(prob),
                   x = dplyr::nth(x, 1), y = dplyr::nth(y, 1),
                   xend = dplyr::nth(x, 2), yend = dplyr::nth(y, 2),
                   .groups = "drop") else NULL

#3.2.3d Self-persistence (diagonal) annotated just outside each node
self_p <- trans |> dplyr::filter(from == to) |>
  dplyr::select(state = from, p_self = prob) |>
  dplyr::right_join(nodes, by = "state")

net <- ggplot()
if (!is.null(edge_paths)) {
  net <- net +
    geom_path(data = edge_paths,
              aes(x, y, group = id, linewidth = prob, alpha = prob),
              colour = "grey35", lineend = "round") +
    geom_segment(data = heads,
                 aes(x = x, y = y, xend = xend, yend = yend, alpha = prob),
                 arrow = grid::arrow(length = grid::unit(0.16, "cm"), type = "closed"),
                 colour = "grey35", linewidth = 0.4, show.legend = FALSE)
}
net +
  geom_point(data = nodes, aes(x, y, size = occupancy, fill = si),
             shape = 21, colour = "grey20", stroke = 0.8) +
  geom_text(data = nodes, aes(x, y, label = state), fontface = "bold", size = 4) +
  geom_text(data = self_p, aes(x * 1.24, y * 1.24,
            label = sprintf("self %.0f%%", 100 * p_self)), size = 2.8, colour = "grey40") +
  scale_linewidth_continuous(range = c(0.3, 2.6), name = "P(transition)") +
  scale_alpha_continuous(range = c(0.25, 0.95), guide = "none") +
  scale_size_continuous(range = c(6, 18), name = "Occupancy (days)") +
  scale_fill_viridis_c(option = "magma", end = 0.9, name = glue::glue("Mean {si_metric_net}")) +
  coord_equal(xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5)) +
  labs(title = "State-transition network",
       subtitle = "Node size = occupancy, fill = mean SI; arrow width = P(move | from); self-% = persistence") +
  theme_void(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 10, colour = "grey40"),
        legend.position = "right")

#3.2.3e Companion heatmap: exact probabilities incl. the diagonal the network
#       abstracts. Network = structure at a glance; heatmap = precise values.
ggplot(trans, aes(factor(to), factor(from), fill = prob)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = ifelse(n == 0, "", sprintf("%.2f", prob))), size = 3) +
  scale_fill_viridis_c(end = 0.95, name = "P(to | from)") +
  labs(title = "Day-to-day state transition probabilities",
       subtitle = "Consecutive study-day pairs only; diagonal = persistence",
       x = "To state (day t+1)", y = "From state (day t)") +
  theme_hw()

```

```{r state_dwell, warning = FALSE}

## 3.2d Dwell times per state ##

#3.2.4 Dwell times: lengths of maximal runs of the SAME state across consecutive
#      days (a run is broken by a state change OR a day gap), i.e. how long the
#      system typically sits in a state before moving.
new_run <- !(seq_df$consecutive & seq_df$state == seq_df$prev_state)
new_run[is.na(new_run)] <- TRUE
seq_df$run <- cumsum(new_run)

dwell <- seq_df |>
  dplyr::group_by(subject, run, state) |>
  dplyr::summarise(dwell = dplyr::n(), .groups = "drop") |>
  dplyr::group_by(state) |>
  dplyr::summarise(n_runs = dplyr::n(),
                   mean_dwell = round(mean(dwell), 2),
                   max_dwell  = max(dwell), .groups = "drop")

nice_tbl(dwell, caption = "Dwell time per state (lengths of consecutive-day runs)")

```

### How to read the critical-slowing-down (CSD) analysis

The theory (Scheffer): a system nearing a tipping point loses resilience -- after
any small perturbation it drifts back to baseline more *slowly*. In daily data
that loss of resilience leaves two statistical fingerprints in the days just
before a transition: (1) **rising variance** (the system wanders further before
recovering) and (2) **rising lag-1 autocorrelation** (today looks more like
yesterday, because recovery is sluggish). The analysis below therefore takes
every stretch of consecutive days, computes those two fingerprints per feature
inside a short trailing window, and asks one question: *are the windows that
immediately precede a within-person SI surge different from all the other
windows?* In the boxplot, each facet column is a feature, the top row is the
autocorrelation fingerprint and the bottom row the variance fingerprint, and
within each panel the two boxes compare "pre-surge" windows against "other"
windows. CSD predicts the pre-surge box sits **higher** in *both* rows. With few
pre-surge windows this is direction-reading only -- preliminary evidence for the
F31, not a test.

```{r early_warning, fig.height = 5, warning = FALSE}

## 3.3 Stage 3b (EXPLORATORY): critical-slowing-down before high-SI onset ##

# WHAT THIS ASKS: dynamical-systems theory (Scheffer) predicts that as a system
# approaches a tipping point, it recovers more slowly from perturbations -- which
# shows up as RISING variance and RISING lag-1 autocorrelation (AC1) in the days
# just before the transition. Here the "transition" is a within-person surge in
# suicidal ideation. If CSD operates at the daily timescale, the short window of
# days immediately BEFORE an SI surge should show higher variance / AC1 in the
# mood+sleep features than ordinary windows. This is the core F31 hypothesis,
# tested here only as a feasibility probe.

si_metric <- if ("si_bother" %in% names(model_df)) "si_bother" else cfg$overlay$primary[1]
ew_window <- 3L   # trailing consecutive-day window (kept short; series are brief/gappy)

#3.3.1 Define ONSET as a within-person SI SURGE rather than a categorical state
#      entry. Categorical entry into the high-SI state is too rare in these short,
#      gappy series to yield usable windows; a within-person rise in the
#      continuous SI score past that person's own median (from below it) captures
#      the same "crossing into a high-SI day" idea with far more events.
si_by_person <- model_df |>
  dplyr::mutate(state_i = as.integer(as.character(state))) |>
  dplyr::group_by(subject) |>
  dplyr::mutate(si_thresh = stats::median(.data[[si_metric]], na.rm = TRUE)) |>
  dplyr::ungroup()
seq_si <- seq_df |>
  dplyr::left_join(dplyr::select(si_by_person, subject, day, si_thresh,
                                 si_now = dplyr::all_of(si_metric)),
                   by = c("subject", "day"))

#3.3.2 Slide a window of `ew_window` CONSECUTIVE days within each subject; compute
#      each feature's variance and AC1 in the window, then flag whether the day
#      immediately AFTER is an SI surge (at/above the person's median, having been
#      below it on the window's last day).
ac1 <- function(x) {
  if (length(x) < 3L || stats::sd(x) == 0) return(NA_real_)
  stats::cor(x[-length(x)], x[-1])
}

ew_rows <- list(); r <- 0L; n_windows <- 0L; n_onset <- 0L
for (sdf in split(seq_si, seq_si$subject)) {
  sdf <- sdf[order(sdf$day), ]
  if (nrow(sdf) < ew_window + 1L) next
  for (i in seq_len(nrow(sdf) - ew_window)) {
    win <- sdf[i:(i + ew_window - 1L), ]
    nxt <- sdf[i + ew_window, ]
    #3.3.2.1 window fully consecutive AND nxt adjacent to its last day
    if (any(diff(win$day) != 1L) || (nxt$day - win$day[ew_window]) != 1L) next
    #3.3.2.2 need SI present on the boundary days to define a surge
    last_si <- win$si_now[ew_window]
    if (is.na(last_si) || is.na(nxt$si_now) || is.na(nxt$si_thresh)) next
    n_windows <- n_windows + 1L
    onset <- (last_si < nxt$si_thresh) && (nxt$si_now >= nxt$si_thresh)
    if (onset) n_onset <- n_onset + 1L
    for (f in feat_model) {
      r <- r + 1L
      ew_rows[[r]] <- data.frame(feature = f,
                                 variance = stats::var(win[[f]]),
                                 ac1 = ac1(win[[f]]), onset = onset)
    }
  }
}
ew <- if (length(ew_rows)) do.call(rbind, ew_rows) else NULL

#3.3.3 FEASIBILITY FIRST: report how many windows and how many pre-onset windows
#      exist. This count -- not any p-value -- is the headline; if pre-onset
#      windows are in the single digits, the boxplots are illustrative only.
message(glue::glue(
  "Early-warning feasibility: {n_windows} valid {ew_window}-day windows, ",
  "of which {n_onset} precede a within-person SI surge ",
  "(SI metric = {si_metric})."))

#3.3.4 Compare indicator levels in pre-surge vs other windows. Strictly
#      hypothesis-generating: read DIRECTION (are pre-surge variance/AC1 higher?)
#      and the pre-surge n, never significance.
ew_ok <- !is.null(ew) && n_onset >= 3L
if (ew_ok) {
  ew_long <- ew |>
    tidyr::pivot_longer(c(variance, ac1), names_to = "indicator", values_to = "value") |>
    tidyr::drop_na(value) |>
    dplyr::mutate(window = ifelse(onset, "pre-surge", "other"),
                  indicator = dplyr::recode(indicator,
                    variance = "variance (within-window spread)",
                    ac1 = "lag-1 autocorrelation (recovery slowing)"))
} else {
  message("Too few pre-surge windows (", n_onset,
          ") for a meaningful comparison; reporting feasibility only. ",
          "This is itself an F31 result: daily sampling is too sparse to resolve ",
          "critical slowing down, motivating denser (e.g. multi-daily) EMA.")
}

#3.3.5 Table + plot are emitted as TOP-LEVEL single-expression conditionals:
#      values created inside a multi-expression `{}` block are invisible to
#      knitr's auto-printing (the summary table silently vanished from an
#      earlier knit for exactly this reason).
if (ew_ok) nice_tbl(
  ew_long |>
    dplyr::group_by(feature, indicator, window) |>
    dplyr::summarise(n = dplyr::n(), mean = mean(value), .groups = "drop"),
  digits = 3,
  caption = glue::glue("Early-warning indicators: pre-surge vs other windows ",
                       "({ew_window}-day windows; {n_onset} pre-surge windows)"))

if (ew_ok) ggplot(ew_long, aes(window, value, fill = window)) +
  geom_boxplot(outlier.size = 0.6, width = 0.6) +
  geom_jitter(width = 0.12, size = 0.5, alpha = 0.3) +
  facet_grid(indicator ~ feature, scales = "free_y", switch = "y",
             labeller = ggplot2::label_wrap_gen(width = 18)) +
  scale_fill_viridis_d(end = 0.7, guide = "none") +
  labs(title = "Critical-slowing-down indicators before a within-person SI surge",
       subtitle = glue::glue("Exploratory; {ew_window}-day windows; ",
                             "CSD predicts pre-surge > other for BOTH rows"),
       x = NULL, y = NULL) +
  theme_hw() +
  theme(strip.placement = "outside", strip.text.x = element_text(size = 8))

```

## Limitations

-   Person-days are not independent (i.e., within-person
    autocorrelation; clustering in \~30 people); the GMM likelihood
    treats them as exchangeable, which inflates the effective N that BIC
    assumes and biases it toward more components. Mitigations: ICL
    (entropy-penalized) and the smallest-mixing-weight diagnostic
    alongside BIC, the person x state artifact check, and the
    person-mean-centered sensitivity refit (Section 2.8) isolating
    within-person state structure.
-   Commission-error rate is excluded from the mixture (documented
    above: \~45% zero-error days force degenerate components) and
    reported as an overlay only.
-   The daily EMA mood items are depression and anxiety only, there is
    no daily elevated-mood/manic item, so states characterize the
    depressive/dysregulated axis, with the manic pole captured only
    indirectly (e.g., via sleep loss).
-   Actigraphy was processed with the Sadeh algorithm for most but not
    all participants; the algorithm is a one-line config swap for a
    Cole-Kripke sensitivity analysis.
-   If the criteria prefer K = 1 or states reduce to person identity,
    that is itself a reportable finding about the (lack of) shared
    multivariate state structure.
-   The Stage-3b early-warning (critical-slowing-down) analysis is
    exploratory and underpowered: short, gap-interrupted per-person
    series yield few consecutive pre-onset windows, so it is reported as
    hypothesis-generating preliminary evidence for the F31, not a
    confirmatory test of rising variance / autocorrelation before
    high-SI transitions.

