---
title: "BBRF GMM: Latent Daily Psychophysiological States & Suicide Risk"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
params:
  profile: "default"
editor_options: 
  markdown: 
    wrap: 72
---

```{r global, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE, comment = "")
```

```{r environment, warning = FALSE, message = FALSE, echo = FALSE}

## Load packages, source code, and config ##

suppressPackageStartupMessages({
  library(here);
  library(dplyr);
  library(readr);
  library(tidyr);
  library(ggplot2);
  library(purrr);
  library(glue);
  library(kableExtra)
})

#0.1 Source the from-scratch model code and shared utilities
source(here::here("src", "utils.R"))
source(here::here("src", "gmm_em.R"))

#0.2 Load config (single source of truth) and set the seed
cfg <- load_config(profile = params$profile)
set.seed(cfg$gmm$seed)

#0.3 Read the analytic person-day table built by R/01_build_dataset.R
daily_path <- list.files(cfg$paths$data_processed, pattern = "^ema_daily",
                         full.names = TRUE) |> sort() |> dplyr::last()

daily <- readr::read_csv(daily_path, show_col_types = FALSE)

```

## Analysis Overview

Goal is to discover latent daily states from a small set of continuous
psychophysiological features (i.e., mood, actigraphy derived sleep,
go/no-go cognition) by fitting a Gaussian Mixture Model via EM, then
characterize those states by their suicidal-ideation profiles and
analyze how individuals move among them over time.

**This will occur in three stages.**

1.  *Univariate EM (EDA + code stepping-stone).* Fit a 1-D GMM to each
    feature to inspect multimodality before committing to the joint
    model. We also run a pre-registered check on the go/no-go commission
    feature and document its eventual exclusion from the mixture
2.  *Multivariate GMM (the model).* Fit the joint mixture; select K by
    comparing BIC, AIC, and ICL (with classification entropy and the
    smallest mixing weight as degeneracy diagnostics), a Kneedle knee, a
    marginal density-fit comparison, and the person x state artifact
    check (never a single minimum)
3.  *States + dynamics.* Overlay SI/SB onto states; decode per-person
    state sequences; relate early-warning indicators (variance, lag-1
    autocorrelation) to transitions into high-SI states

```{r prepare_features, echo = FALSE, warning = FALSE}

## 1. Build the standardized feature matrix (complete cases) ##

#1.1 Configured features. The FULL set (`feat`) drives Stage-1 EDA and the
#    documented commission check; the MODELING set (`feat_model`) drops any
#    features flagged for documented exclusion (e.g., commission_elogit) so they
#    never enter the mixture, while their evidence is retained.
feat <- cfg$features
feat_excl <- if (!is.null(cfg$gmm$exclude_from_mixture)) unlist(cfg$gmm$exclude_from_mixture) else character(0)
feat_model <- setdiff(feat, feat_excl)
stopifnot(all(feat %in% names(daily)))

#1.2 Complete-case on the MODELING features (so an excluded feature's missingness
#    can't shrink N), while carrying the excluded feature(s) + overlays for
#    documentation and Stage-3 characterization.
model_df <- daily |>
  dplyr::select(subject, day, dplyr::all_of(feat),
                dplyr::any_of(c(cfg$overlay$primary, cfg$overlay$secondary, cfg$overlay$rare))) |>
  tidyr::drop_na(dplyr::all_of(feat_model))

message(glue::glue(
  "Modeling on {nrow(model_df)} complete person-days ",
  "across {dplyr::n_distinct(model_df$subject)} subjects ",
  "(of {nrow(daily)} total person-days). Modeling features: ",
  "{paste(feat_model, collapse = ', ')}."))

#1.3 Standardize the MODELING features; keep person id aligned for the artifact check
X <- zscore_matrix(model_df[, feat_model])
person_id <- model_df$subject

```

## 1. Univariate EM (per-feature multimodality)

```{r stage1_feat_dists, echo = FALSE, warning = FALSE}

## Stage 1a: feature distributions (shape QC) shown for the FULL set ##

#1a.1 Long-format the full feature set in native units, incl. commission so its
#     floor-heavy distribution is documented before the exclusion section
feat_long <- model_df |>
  dplyr::select(dplyr::all_of(feat)) |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "feature", values_to = "value") |>
  tidyr::drop_na(value)

#1a.2 Faceted histograms: read for skew, spikes/floors, and modality
ggplot(feat_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "#21908C", colour = "white", linewidth = 0.2) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  labs(title = "Feature distributions (native units)", x = NULL, y = "Person-days") +
  theme_hw()

#1a.3 Numeric shape summary, incl. share sitting exactly at the floor (the spike)
feat_long |>
  dplyr::group_by(feature) |>
  dplyr::summarise(
    n = dplyr::n(),
    mean = mean(value), sd = stats::sd(value),
    skew = mean(((value - mean(value)) / stats::sd(value))^3),
    pct_at_floor = round(100 * mean(value == min(value)), 1),
    .groups  = "drop") |>
  nice_tbl(caption = "Feature shape summary (skew + floor share)")

```

```{r stage1_univariate_em, warning = FALSE}

## Fit a 2-component 1-D GMM to each feature as exploratory structure check ##

uni_fits <- purrr::map(feat, function(v) {
  x <- model_df[[v]]
  x <- x[is.finite(x)]
  gmm_em_fit_1d(x, K = 2L, n_restarts = cfg$gmm$n_restarts, seed = cfg$gmm$seed)
})

names(uni_fits) <- feat

#1.1 Summarize each feature's recovered component means / weights
uni_summary <- purrr::imap_dfr(uni_fits, function(f, nm) {
  ord <- order(f$mu)
  tibble::tibble(
    feature = nm,
    mu_low = f$mu[ord][1], mu_high = f$mu[ord][2],
    sd_low = sqrt(f$sigma2[ord][1]), sd_high = sqrt(f$sigma2[ord][2]),
    w_low = f$weights[ord][1], w_high  = f$weights[ord][2],
    separation = abs(diff(f$mu)) / sqrt(mean(f$sigma2)))
})

nice_tbl(uni_summary, caption = "Univariate 2-component summaries", digits = 3)

```

## Commission: documented exclusion check

Go/no-go commission-error rate is a theoretically central index of
inhibitory control, and we retained it as a candidate state feature. But
\~45% of person-days have zero commission errors (\<=1 error on \~50
trials; see the floor share above), so the empirical-logit rate carries
a point mass that a Gaussian component can only absorb by collapsing its
variance. The check below fits the mixture with commission included and
shows the consequence, i.e., commission dominates feature discrimination
only by spawning near-zero-variance components anchored at the
error-rate floor. In other words, it sorts days into "any errors vs
none" rather than indexing graded control. On that pre-registered basis
we exclude it from the mixture and retain it as a descriptive overlay on
the mood/sleep states. The evidence is kept here for the writeup.

```{r commission_documented_exclusion, warning = FALSE}

## Why commission is an overlay, not a mixture input ##

if ("commission_elogit" %in% feat) {

  #C.1 Fit on commission-complete days, on the FULL feature set, at a documented K
  doc_rows <- which(stats::complete.cases(model_df[, feat]))
  X_doc <- zscore_matrix(model_df[doc_rows, feat])
  k_doc <- cfg$gmm$commission_doc_k

  fit_doc <- gmm_em_fit(X_doc, K = k_doc, n_restarts = cfg$gmm$n_restarts,
                        max_iter = cfg$gmm$max_iter, tol = cfg$gmm$tol,
                        ridge = cfg$gmm$ridge, seed = cfg$gmm$seed)

  #C.2 Commission dominates discrimination (via near-zero within-component variance)
  nice_tbl(gmm_feature_discrimination(fit_doc, X_doc),
           caption = glue::glue("Commission INCLUDED (K = {k_doc}): per-feature discrimination ",
                               "(commission's ratio is inflated by a near-zero-variance spike)"))

  #C.3 The smoking gun: states whose commission var_z sits at the ridge are the
  #    degenerate zero-error spike(s)
  ci <- which(colnames(X_doc) == "commission_elogit")
  nice_tbl(tibble::tibble(
    state = factor(seq_len(fit_doc$K)),
    mean_z = round(fit_doc$mu[, ci], 2),
    var_z  = round(vapply(fit_doc$Sigma, function(S) S[ci, ci], numeric(1)), 4),
    weight = round(fit_doc$weights, 3)),
    caption = "commission_elogit by state (var_z near the GMM ridge => degenerate spike)")

} else {
  message("commission_elogit not in the configured feature set; skipping documented check.")
}

```

## 2. Multivariate GMM + model selection

```{r gmm_select_k, warning = FALSE}

## Fit across K; compare BIC, AIC, ICL (+ entropy, smallest mixing weight) ##

sel <- gmm_select_k(
  X,
  k_range = unlist(cfg$gmm$k_range),
  n_restarts = cfg$gmm$n_restarts,
  max_iter = cfg$gmm$max_iter,
  tol = cfg$gmm$tol,
  ridge = cfg$gmm$ridge,
  seed = cfg$gmm$seed)

nice_tbl(sel$bic_table,
         caption = "Model selection across K (BIC, AIC, ICL, entropy, min mixing weight)",
         digits = 1)

#2.1 Objective knees on the BIC and ICL curves
k_knee_bic <- kneedle(sel$bic_table$K, sel$bic_table$bic)
k_knee_icl <- kneedle(sel$bic_table$K, sel$bic_table$icl)

message(glue::glue("BIC min K={sel$best_k} | ICL min K={sel$best_k_icl} | ",
                   "Kneedle(BIC) K={k_knee_bic} | Kneedle(ICL) K={k_knee_icl}"))

#2.2 BIC vs ICL: ICL adds an overlap/entropy penalty, so it resists the
#    over-splitting BIC shows when the mixture tiles density rather than finding
#    distinct subgroups
crit_long <- sel$bic_table |>
  dplyr::select(K, BIC = bic, ICL = icl) |>
  tidyr::pivot_longer(c(BIC, ICL), names_to = "criterion", values_to = "value")

ggplot(crit_long, aes(K, value, colour = criterion)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.4) +
  geom_vline(xintercept = sel$best_k_icl, linetype = 2, colour = "grey50") +
  scale_colour_viridis_d(end = 0.6) +
  scale_x_continuous(breaks = sel$bic_table$K) +
  labs(title = "Model selection: BIC vs ICL",
       subtitle = glue::glue("dashed = ICL-min (K={sel$best_k_icl}); ",
                             "Kneedle(ICL) K={k_knee_icl}; lower is better"),
       x = "Number of states (K)", y = "Criterion") +
  theme_hw()

#2.3 Candidate K set for the density comparison (DYNAMIC; config-overridable).
#    Defaults to the ICL minimum, the ICL knee, and K=2 if not specified given
#    K=2 relevance derived from viewing the density plots
candidate_k <- if (!is.null(cfg$gmm$candidate_k)) {
  sort(unique(unlist(cfg$gmm$candidate_k)))
} else {
  sort(unique(c(2L, sel$best_k_icl, k_knee_icl)))
}

```

```{r gmm_choose_k, warning = FALSE, echo = FALSE}

## 2.4 Lock K from all lenses (config override, else the ICL minimum) ##
chosen_k <- if (!is.null(cfg$gmm$chosen_k)) cfg$gmm$chosen_k else sel$best_k_icl
best <- sel$fits[[as.character(chosen_k)]]
message(glue::glue(
  "Carrying K = {chosen_k} ",
  "({ifelse(is.null(cfg$gmm$chosen_k), 'ICL default', 'config override')}). ",
  "Set cfg$gmm$chosen_k in config to lock it after reviewing the criteria table + curve."))

```

```{r gmm_density_compare, fig.height = 7, warning = FALSE}

## 2.5 What each state's distribution looks like (at the chosen K) ##

#2.5.1 PLOT A: per-feature component densities, one curve PER STATE, over the
#      data histogram. Reads as: "for this feature, here is each state's density
#      and where it sits." The summed (mixture) density is the dashed envelope
comp_dens <- purrr::map_dfr(colnames(X),
  function(f) gmm_component_density(best, X, f))

mix_dens  <- purrr::map_dfr(colnames(X),
  function(f) gmm_marginal_density(best, X, f))

model_long <- model_df |>
  dplyr::select(dplyr::all_of(colnames(X))) |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "feature", values_to = "value")

ggplot() +
  geom_histogram(data = model_long, aes(value, y = ggplot2::after_stat(density)),
                 bins = 30, fill = "grey88", colour = "white", linewidth = 0.2) +
  geom_area(data = comp_dens, aes(value, density, fill = state),
            position = "identity", alpha = 0.35) +
  geom_line(data = comp_dens, aes(value, density, colour = state), linewidth = 0.7) +
  geom_line(data = mix_dens, aes(value, density), colour = "grey25",
            linetype = 2, linewidth = 0.5) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  scale_fill_viridis_d(end = 0.9) + scale_colour_viridis_d(end = 0.9) +
  labs(title = glue::glue("State distributions by feature (K = {best$K})"),
       subtitle = "Each colour = one state's weighted Gaussian; dashed grey = full mixture",
       x = NULL, y = "Density") +
  theme_hw()

#2.5.2 PLOT B: the mood plane, coloured by hard state assignment, with each
#      state's mean marked. Depression x anxiety = the dominant axis
if (all(c("depression", "anxiety") %in% colnames(X))) {
  state_means <- gmm_state_profile(best, X)

  ggplot(model_df, aes(depression, anxiety)) +
    geom_point(aes(colour = factor(gmm_map_assign(best))), alpha = 0.5, size = 1.4) +
    geom_point(data = state_means, aes(depression, anxiety),
               shape = 21, size = 5, stroke = 1.1, fill = "white", colour = "grey20") +
    geom_text(data = state_means, aes(depression, anxiety, label = state),
              size = 3.2, fontface = "bold") +
    scale_colour_viridis_d(end = 0.9, name = "State") +
    labs(title = glue::glue("Person-days on the mood plane, by state (K = {best$K})"),
         subtitle = "Points = person-days (coloured by MAP state); numbered rings = state means",
         x = "Depression", y = "Anxiety") +
    theme_hw()
}

#2.5.3 PLOT C: standardized state x feature heatmap (compact fingerprint).
#      A single grid: rows = states, columns = features, fill = z-mean. Fast read
#      of "what is high/low in each state" across ALL features at once
fing_long <- best$mu |>
  `colnames<-`(colnames(X)) |>
  tibble::as_tibble() |>
  dplyr::mutate(state = factor(seq_len(best$K))) |>
  tidyr::pivot_longer(-state, names_to = "feature", values_to = "z_mean")

ggplot(fing_long, aes(feature, state, fill = z_mean)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%+.1f", z_mean)), size = 3) +
  scale_fill_gradient2(low = "#3B528B", mid = "grey95", high = "#B8003E",
                       midpoint = 0, name = "z-mean") +
  labs(title = glue::glue("State fingerprints (standardized; K = {best$K})"),
       subtitle = "Fill = standardized mean (0 = sample average); +red above, -blue below",
       x = NULL, y = "State") +
  theme_hw() + theme(axis.text.x = element_text(angle = 30, hjust = 1))

```

```{r gmm_state_profiles, warning = FALSE}

## 2.6 What ARE the states? Native-unit profiles + standardized fingerprints ##

nice_tbl(gmm_state_profile(best, X), digits = 2,
         caption = "Latent state profiles (native units) and state size")

#2.6.1 Standardized fingerprint (z-units; 0 = sample average), lollipop per state
fing <- best$mu

colnames(fing) <- colnames(X)

fing <- tibble::as_tibble(fing)

fing$state <- factor(seq_len(best$K))

fing_long <- tidyr::pivot_longer(fing, -state, names_to = "feature", values_to = "z_mean")

ggplot(fing_long, aes(x = z_mean, y = feature, colour = state)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey60") +
  geom_segment(aes(x = 0, xend = z_mean, yend = feature), colour = "grey80", linewidth = 0.6) +
  geom_point(size = 3) +
  facet_wrap(~ state, nrow = 1) +
  scale_colour_viridis_d(end = 0.9) +
  labs(title = "Latent state fingerprints",
       subtitle = "Standardized feature means by state (0 = sample average)",
       x = "Standardized mean (SD units)", y = NULL) +
  theme_hw() + theme(legend.position = "none")

```

```{r gmm_artifact_check, warning = FALSE}

## 2.7 Person x state check: is any state just one person's days? ##
chk <- gmm_person_state_check(best, person_id)

nice_tbl(data.frame(state = seq_along(chk$max_person_share),
                    n_days = as.integer(chk$state_totals),
                    max_single_person_share = round(chk$max_person_share, 3)),
         caption = "Per-state size and single-person dominance (high share => trait artifact)")

```

## 3. State characterization, SI overlay, and dynamics

```{r state_overlay, warning = FALSE}

## 3.1 MAP-assign states and overlay SI prevalence (+ commission) per state ##

model_df$state <- factor(gmm_map_assign(best))

overlay_vars <- intersect(c(cfg$overlay$primary, cfg$overlay$secondary, cfg$overlay$rare),
                          names(model_df))
state_overlay <- model_df |>
  dplyr::group_by(state) |>
  dplyr::summarise(
    n_days = dplyr::n(),
    dplyr::across(dplyr::all_of(overlay_vars), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop")

nice_tbl(state_overlay, digits = 3, caption = "SI/SB (and commission overlay) profile by latent state")

```

```{r dynamics, eval = FALSE, warning = FALSE}

## 3.2 Dynamics (Stage 3): per-person state sequences + early-warning signals ##
# TODO: for each subject, order states by `day`, then compute
#   - dwell times and transition frequencies among states
#   - rolling variance and lag-1 autocorrelation of each feature in the window
#     preceding transitions INTO high-SI states (critical slowing down, Scheffer)
# This is where the trajectory/dynamical-systems story is recovered from the
# (otherwise day-exchangeable) Gaussian mixture.

```

## Limitations

-   Person-days are not independent (i.e., within-person
    autocorrelation; clustering in \~30 people); the GMM likelihood
    treats them as exchangeable, which inflates the effective N that BIC
    assumes and biases it toward more components. Mitigations: ICL
    (entropy-penalized) and the smallest-mixing-weight diagnostic
    alongside BIC, the person x state artifact check, and a planned
    person-mean-centered fit (`person_center()`) isolating within-person
    state structure.
-   Commission-error rate is excluded from the mixture (documented
    above: \~45% zero-error days force degenerate components) and
    reported as an overlay only.
-   The daily EMA mood items are depression and anxiety only -- there is
    no daily elevated-mood/manic item -- so states characterize the
    depressive/dysregulated axis, with the manic pole captured only
    indirectly (e.g., via sleep loss).
-   Actigraphy was processed with the Sadeh algorithm for most but not
    all participants; the algorithm is a one-line config swap for a
    Cole-Kripke sensitivity analysis.
-   If the criteria prefer K = 1 or states reduce to person identity,
    that is itself a reportable finding about the (lack of) shared
    multivariate state structure.
