---
title: "BBRF GMM: Latent Daily Psychophysiological States & Suicide Risk"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
params:
  profile: "default"
---

```{r global, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE, comment = "")
```

```{r environment, message = FALSE}

## Load packages, source code, and config ##
suppressPackageStartupMessages({
  library(here);
  library(dplyr);
  library(readr);
  library(tidyr);
  library(ggplot2);
  library(purrr);
  library(glue)
})

#0.1 Source the from-scratch model code and shared utilities
source(here::here("R", "utils.R"))
source(here::here("R", "gmm_em.R"))

#0.2 Load config (single source of truth) and set the seed
cfg <- load_config(profile = params$profile)
set.seed(cfg$gmm$seed)

#0.3 Read the analytic person-day table built by R/01_build_dataset.R
daily_path <- list.files(cfg$paths$data_processed, pattern = "^ema_daily",
                         full.names = TRUE) |> sort() |> dplyr::last()

daily <- readr::read_csv(daily_path, show_col_types = FALSE)

```

## Analysis Overview

Goal is to discover latent daily states from a small set of continuous
psychophysiological features (i.e., mood, actigraphy derived sleep, go/no-go cognition) by
fitting a Gaussian Mixture Model via EM, then characterize those states by their
suicidal-ideation profiles and analyze how individuals move among them over time.

**This will occur in three stages.**

1. *Univariate EM (EDA + code stepping-stone).* Fit a 1-D GMM to each feature to
   inspect multimodality before committing to the joint model
2. *Multivariate GMM (the model).* Fit the joint mixture, select K by BIC with
   multi-restart stability, and guard against person-identity artifacts
3. *States + dynamics.* Overlay SI/SB onto states; decode per-person state
   sequences; relate early-warning indicators (variance, lag-1 autocorrelation)
   to transitions into high-SI states

```{r prepare-features}

## 1. Build the standardized feature matrix (complete cases) ##

#1.1 Pull the configured GMM input features
feat <- cfg$features
stopifnot(all(feat %in% names(daily)))

#1.2 Complete-case person-days across the feature set (report the retained N)
model_df <- daily |>
  dplyr::select(subject, day, dplyr::all_of(feat),
                dplyr::any_of(c(cfg$overlay$primary, cfg$overlay$secondary, cfg$overlay$rare))) |>
  tidyr::drop_na(dplyr::all_of(feat))

message(glue::glue(
  "Modeling on {nrow(model_df)} complete person-days ",
  "across {dplyr::n_distinct(model_df$subject)} subjects ",
  "(of {nrow(daily)} total person-days)."
))

#1.3 Standardize features (z-score); keep person id aligned for the artifact check
X <- zscore_matrix(model_df[, feat])
person_id <- model_df$subject

```

## 1. Univariate EM (per-feature multimodality)

```{r univariate-em}

## Fit a 2-component 1-D GMM to each feature as exploratory structure check ##

uni_fits <- purrr::map(feat, function(v) {
  x <- model_df[[v]]
  x <- x[is.finite(x)]
  gmm_em_fit_1d(x, K = 2L, n_restarts = cfg$gmm$n_restarts, seed = cfg$gmm$seed)
})
names(uni_fits) <- feat

#1.1 Summarize each feature's recovered component means / weights
uni_summary <- purrr::imap_dfr(uni_fits, function(f, nm) {
  data.frame(feature = nm,
             mu_1 = sort(f$mu)[1], mu_2 = sort(f$mu)[2],
             weight_min = min(f$weights))
})

knitr::kable(uni_summary, digits = 3, caption = "Univariate 2-component summaries")

```

## 2. Multivariate GMM + model selection

```{r select-k}

## Fit across K and select by BIC (with multi-restart stability) ##

sel <- gmm_select_k(
  X,
  k_range = unlist(cfg$gmm$k_range),
  n_restarts = cfg$gmm$n_restarts,
  max_iter = cfg$gmm$max_iter,
  tol = cfg$gmm$tol,
  ridge = cfg$gmm$ridge,
  seed = cfg$gmm$seed)

knitr::kable(sel$bic_table, digits = 1, caption = "Model selection across K (BIC)")
message("BIC-selected K = ", sel$best_k)

#2.1 BIC curve (decluttered)
ggplot(sel$bic_table, aes(x = K, y = bic)) +
  geom_line(colour = "grey60") +
  geom_point(size = 2) +
  geom_point(data = subset(sel$bic_table, K == sel$best_k),
             colour = state_palette(1), size = 3.5) +
  labs(title = "BIC by number of states", x = "K (states)", y = "BIC") +
  theme_swd()

best <- sel$best_fit

```

```{r artifact-check}

## 2.2 Person x state check: is any state just one person's days? ##

chk <- gmm_person_state_check(best, person_id)
art <- data.frame(state = seq_along(chk$max_person_share),
                  n_days = as.integer(chk$state_totals),
                  max_single_person_share = round(chk$max_person_share, 3))

knitr::kable(art, caption = "Per-state size and single-person dominance (high share => artifact)")

```

## 3. State characterization, SI overlay, and dynamics

```{r state-overlay}

## 3.1 MAP-assign states and overlay SI prevalence per state ##

model_df$state <- factor(gmm_map_assign(best))

overlay_vars <- intersect(c(cfg$overlay$primary, cfg$overlay$secondary, cfg$overlay$rare),
                          names(model_df))
state_profile <- model_df |>
  dplyr::group_by(state) |>
  dplyr::summarise(
    n_days = dplyr::n(),
    dplyr::across(dplyr::all_of(overlay_vars), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop")

knitr::kable(state_profile, digits = 3, caption = "SI/SB profile by latent state")

```

```{r dynamics, eval = FALSE}

## 3.2 Dynamics (Stage 3): per-person state sequences + early-warning signals ##
# TODO (needs real data): for each subject, order states by `day`, then compute
#   - dwell times and transition frequencies among states
#   - rolling variance and lag-1 autocorrelation of each feature in the window
#     preceding transitions INTO high-SI states (critical slowing down, Scheffer)
# This is where the trajectory/dynamical-systems story is recovered from the
# (otherwise day-exchangeable) Gaussian mixture.

```

## Limitations

- Person-days are not independent (within-person autocorrelation; clustering in
  ~32 people); the GMM likelihood treats them as exchangeable. Mitigations: the
  person x state artifact check above, and a planned person-mean-centered fit
  (`person_center()`) isolating within-person state structure.
- Actigraphy was processed with the Sadeh algorithm for most but not all
  participants; the algorithm is a one-line config swap for a Cole-Kripke
  sensitivity analysis.
- If BIC prefers K = 1 or states reduce to person identity, that is itself a
  reportable finding about the (lack of) shared multivariate state structure.
