---
title: "BBRF GMM: Latent Daily Psychophysiological States & Suicide Risk"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
params:
  profile: "default"
---

```{r global, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE, comment = "")
```

```{r environment, message = FALSE}

## Load packages, source code, and config ##
suppressPackageStartupMessages({
  library(here);
  library(dplyr);
  library(readr);
  library(tidyr);
  library(ggplot2);
  library(purrr);
  library(glue);
  library(kableExtra)
})

#0.1 Source the from-scratch model code and shared utilities
source(here::here("src", "utils.R"))
source(here::here("src", "gmm_em.R"))

#0.2 Load config (single source of truth) and set the seed
cfg <- load_config(profile = params$profile)
set.seed(cfg$gmm$seed)

#0.3 Read the analytic person-day table built by R/01_build_dataset.R
daily_path <- list.files(cfg$paths$data_processed, pattern = "^ema_daily",
                         full.names = TRUE) |> sort() |> dplyr::last()

daily <- readr::read_csv(daily_path, show_col_types = FALSE)

```

## Analysis Overview

Goal is to discover latent daily states from a small set of continuous
psychophysiological features (i.e., mood, actigraphy derived sleep, go/no-go cognition) by
fitting a Gaussian Mixture Model via EM, then characterize those states by their
suicidal-ideation profiles and analyze how individuals move among them over time.

**This will occur in three stages.**

1. *Univariate EM (EDA + code stepping-stone).* Fit a 1-D GMM to each feature to
   inspect multimodality before committing to the joint model
2. *Multivariate GMM (the model).* Fit the joint mixture, select K by BIC with
   multi-restart stability, and guard against person-identity artifacts
3. *States + dynamics.* Overlay SI/SB onto states; decode per-person state
   sequences; relate early-warning indicators (variance, lag-1 autocorrelation)
   to transitions into high-SI states

```{r prepare-features, echo = FALSE, warning = FALSE}

## 1. Build the standardized feature matrix (complete cases) ##

#1.1 Pull the configured GMM input features
feat <- cfg$features
stopifnot(all(feat %in% names(daily)))

#1.2 Complete-case person-days across the feature set (report the retained N)
model_df <- daily |>
  dplyr::select(subject, day, dplyr::all_of(feat),
                dplyr::any_of(c(cfg$overlay$primary, cfg$overlay$secondary, cfg$overlay$rare))) |>
  tidyr::drop_na(dplyr::all_of(feat))

message(glue::glue(
  "Modeling on {nrow(model_df)} complete person-days ",
  "across {dplyr::n_distinct(model_df$subject)} subjects ",
  "(of {nrow(daily)} total person-days)."))

#1.3 Standardize features (z-score); keep person id aligned for the artifact check
X <- zscore_matrix(model_df[, feat])
person_id <- model_df$subject

```

## 1. Univariate EM (per-feature multimodality)

```{r feat dists, echo = FALSE, warning = FALSE}

## Stage 1a: feature distributions (shape QC) ##

#1a.1 Long-format the GMM input features in native units
feat_long <- model_df |>
  dplyr::select(dplyr::all_of(feat)) |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "feature", values_to = "value")

#1a.2 Faceted histograms: read for skew, spikes/floors, and modality
ggplot(feat_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "#21908C", colour = "white", linewidth = 0.2) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  labs(title = "Feature distributions (native units)", x = NULL, y = "Person-days") +
  theme_hw()

#1a.3 Numeric shape summary, incl. share sitting exactly at the floor (the spike)
feat_long |>
  dplyr::group_by(feature) |>
  dplyr::summarise(
    n = dplyr::n(),
    mean = mean(value), sd = stats::sd(value),
    skew = mean(((value - mean(value)) / stats::sd(value))^3),
    pct_at_floor = round(100 * mean(value == min(value)), 1),
    .groups  = "drop") |>
  nice_tbl(caption = "Feature shape summary (skew + floor share)")

```


```{r univariate-em}

## Fit a 2-component 1-D GMM to each feature as exploratory structure check ##

uni_fits <- purrr::map(feat, function(v) {
  x <- model_df[[v]]
  x <- x[is.finite(x)]
  gmm_em_fit_1d(x, K = 2L, n_restarts = cfg$gmm$n_restarts, seed = cfg$gmm$seed)
})

names(uni_fits) <- feat

#1.1 Summarize each feature's recovered component means / weights
uni_summary <- purrr::imap_dfr(uni_fits, function(f, nm) {
  ord <- order(f$mu)
  tibble::tibble(
    feature = nm,
    mu_low = f$mu[ord][1], mu_high = f$mu[ord][2],
    sd_low = sqrt(f$sigma2[ord][1]), sd_high = sqrt(f$sigma2[ord][2]),
    w_low = f$weights[ord][1], w_high  = f$weights[ord][2],
    separation = abs(diff(f$mu)) / sqrt(mean(f$sigma2)))
})

nice_tbl(uni_summary, caption = "Univariate 2-component summaries", digits = 3)

```

## 2. Multivariate GMM + model selection

```{r gmm_select_k}

## Fit across K and select by BIC (with multi-restart stability) ##

sel <- gmm_select_k(
  X,
  k_range = unlist(cfg$gmm$k_range),
  n_restarts = cfg$gmm$n_restarts,
  max_iter = cfg$gmm$max_iter,
  tol = cfg$gmm$tol,
  ridge = cfg$gmm$ridge,
  seed = cfg$gmm$seed)

nice_tbl(sel$bic_table, caption = "Model selection across K (BIC)", digits = 1)

message("BIC-selected K = ", sel$best_k)

best <- sel$best_fit

#2.1 BIC curve: lower is better; read the elbow, not just the minimum
ggplot(sel$bic_table, aes(x = K, y = bic)) +
  geom_line(colour = "grey70", linewidth = 0.6) +
  geom_point(size = 2.5, colour = "grey45") +
  geom_point(data = subset(sel$bic_table, K == sel$best_k),
             size = 4, colour = "#21908C") +
  scale_x_continuous(breaks = sel$bic_table$K) +
  labs(title = "Model selection by BIC",
       subtitle = "Lower is better \u2014 read the elbow, not just the minimum",
       x = "Number of states (K)", y = "BIC") +
  theme_hw()

```

```{r gmm_state_profiles}

## 2.2 What ARE the states? Native-unit profiles + standardized fingerprints ##

nice_tbl(gmm_state_profile(best, X), digits = 2,
         caption = "Latent state profiles (native units) and state size")

#2.2.1 Standardized fingerprint (z-units; 0 = sample average), lollipop per state
fing <- best$mu

colnames(fing) <- colnames(X)

fing <- tibble::as_tibble(fing)

fing$state <- factor(seq_len(best$K))

fing_long  <- tidyr::pivot_longer(fing, -state, names_to = "feature", values_to = "z_mean")

ggplot(fing_long, aes(x = z_mean, y = feature, colour = state)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey60") +
  geom_segment(aes(x = 0, xend = z_mean, yend = feature), colour = "grey80", linewidth = 0.6) +
  geom_point(size = 3) +
  facet_wrap(~ state, nrow = 1) +
  scale_colour_viridis_d(end = 0.9) +
  labs(title = "Latent state fingerprints",
       subtitle = "Standardized feature means by state (0 = sample average)",
       x = "Standardized mean (SD units)", y = NULL) +
  theme_hw() + theme(legend.position = "none")

```


```{r gmm_artifact_check}

## 2.2 Person x state check: is any state just one person's days? ##
chk <- gmm_person_state_check(best, person_id)

nice_tbl(data.frame(state = seq_along(chk$max_person_share),
                    n_days = as.integer(chk$state_totals),
                    max_single_person_share = round(chk$max_person_share, 3)),
         caption = "Per-state size and single-person dominance (high share => trait artifact)")

```

```{r gmm_commission_check}

## 2.4 Pre-registered commission check: does it separate states, or just spike? ##

nice_tbl(gmm_feature_discrimination(best, X),
         caption = "Per-feature discrimination (between/within); commission near the bottom => demotion candidate")

#2.4.1 commission_elogit by state: a near-ridge var_z at the floor = degenerate spike
if ("commission_elogit" %in% colnames(X)) {
  ci <- which(colnames(X) == "commission_elogit")
  nice_tbl(tibble::tibble(
    state = factor(seq_len(best$K)),
    mean_z = round(best$mu[, ci], 2),
    var_z = round(vapply(best$Sigma, function(S) S[ci, ci], numeric(1)), 4),
    weight = round(best$weights, 3)),
    caption = "commission_elogit by state (var_z near the GMM ridge => degenerate spike)")
}

```


## 3. State characterization, SI overlay, and dynamics

```{r state-overlay}

## 3.1 MAP-assign states and overlay SI prevalence per state ##

model_df$state <- factor(gmm_map_assign(best))

overlay_vars <- intersect(c(cfg$overlay$primary, cfg$overlay$secondary, cfg$overlay$rare),
                          names(model_df))
state_profile <- model_df |>
  dplyr::group_by(state) |>
  dplyr::summarise(
    n_days = dplyr::n(),
    dplyr::across(dplyr::all_of(overlay_vars), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop")

knitr::kable(state_profile, digits = 3, caption = "SI/SB profile by latent state")

```

```{r dynamics, eval = FALSE}

## 3.2 Dynamics (Stage 3): per-person state sequences + early-warning signals ##
# TODO (needs real data): for each subject, order states by `day`, then compute
#   - dwell times and transition frequencies among states
#   - rolling variance and lag-1 autocorrelation of each feature in the window
#     preceding transitions INTO high-SI states (critical slowing down, Scheffer)
# This is where the trajectory/dynamical-systems story is recovered from the
# (otherwise day-exchangeable) Gaussian mixture.

```

## Limitations

- Person-days are not independent (within-person autocorrelation; clustering in
  ~32 people); the GMM likelihood treats them as exchangeable. Mitigations: the
  person x state artifact check above, and a planned person-mean-centered fit
  (`person_center()`) isolating within-person state structure.
- Actigraphy was processed with the Sadeh algorithm for most but not all
  participants; the algorithm is a one-line config swap for a Cole-Kripke
  sensitivity analysis.
- If BIC prefers K = 1 or states reduce to person identity, that is itself a
  reportable finding about the (lack of) shared multivariate state structure.
